---
title: "IMA Day 1 Notes"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---




## Getting Started

**Load packages required for exercises:**

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(readr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(choroplethr))
suppressPackageStartupMessages(library(choroplethrMaps))
```


<!-- **Load the data** -->

<!-- ```{r} -->
<!-- #this uses the readr library: -->
<!-- elect <- read.csv("https://raw.githubusercontent.com/tonmcg/County_Level_Election_Results_12-16/master/US_County_Level_Presidential_Results_08-16.csv") -->
<!-- ``` -->



<!-- <br> -->
<!-- <br> -->


<!-- ## Pre-Bootcamp Review -->

<!-- ### Exercise 1: Explore the structure of the data -->

<!-- ```{r} -->
<!-- # Check out the first rows of `elect`.  What are the cases?  The variables? -->
<!-- head(elect) -->

<!-- # How much data do we have? -->
<!-- dim(elect) -->

<!-- # What are the names of the variables? -->
<!-- names(elect) -->

<!-- #How can we subset just the `dem_2016` data? -->
<!-- head(elect$dem_2016) -->

<!-- #How can we filter out only the counties in which more than 25,000 people voted in 2016? -->
<!-- library(plyr) -->
<!-- filter(elect, total_2016 > 25000) -->
<!-- ``` -->



<!-- <br> -->
<!-- <br> -->

<!-- ### Exercise 2: Create new variables -->


<!-- ```{r} -->
<!-- #start with 1 variable to get the hang of it -->


<!-- #now 2 variables     -->


<!-- #now 3.  (why do we put winrep_2016 in its own row?!) -->


<!-- #finally, add all 2012 and 2016 variables and store as 'elect' -->
<!-- elect = mutate(.data = elect, f_dem_2016 = dem_2016 / total_2016, f_gop_2016 = gop_2016 / total_2016, win_rep = ) -->

<!-- ``` -->



<!-- <br> -->
<!-- <br> -->

<!-- ### Exercise 3: Explore the win column -->

<!-- ```{r} -->
<!-- #Construct a table (a numerical summary) of the number of counties that Trump won/lost -->

<!-- #Construct a bar chart (a visual summary) of this variable.     -->

<!-- ``` -->





<!-- <br> -->
<!-- <br> -->

<!-- ### Exercise 4: Visually explore vote percentages -->

<!-- ```{r} -->
<!-- #histogram -->

<!-- #density plot -->

<!-- ``` -->



<!-- <br> -->
<!-- <br> -->

<!-- ### Exercise 5: Numerically summarize vote percentages -->

<!-- ```{r} -->
<!-- #calculate the mean Trump support -->

<!-- #calculate the median Trump support -->

<!-- #calculate the variance in Trump support from county to county -->

<!-- #calculate the standard deviation in Trump support from county to county -->

<!-- #use the 68-95-99.7 Rule to approximate the range of from the 2.5th to 97.5th percentiles of  Trump support -->

<!-- #calculate the exact 2.5th and 97.5th percentiles  -->
<!-- ``` -->


<!-- <br> -->
<!-- <br> -->
<!-- <br> -->
<!-- <br> -->


<!-- ## Explaining Variability + Wrangling -->


<!-- **Load the data:** -->

<!-- ```{r warning=FALSE} -->
<!-- #load demographic data from the choroplethr library -->
<!-- data("df_county_demographics") -->
<!-- head(df_county_demographics,3) -->

<!-- #load RedBluePurple data -->
<!-- RedBlue <- read.csv("https://www.macalester.edu/~ajohns24/Data/RedBluePurple.csv") -->
<!-- head(RedBlue,3) -->
<!-- ``` -->



<!-- <br> -->
<!-- <br> -->

<!-- **Join / merge these data tables into a single data set:** -->

<!-- ```{r} -->
<!-- #define "region" variable -->

<!-- #join elect and df_county_demographics (from dplyr) -->

<!-- #join the new fulldata with RedBlue -->

<!-- #check out the names of fulldata -->
<!-- ``` -->

```{r Load Data}
fulldata = read.csv("https://www.macalester.edu/~ajohns24/data/IMAdata1.csv")
```


<br>
<br>
<br>
<br>


## Visualizing Relationships

### Exercise 6: Scatterplots of 2 quantitative variables


```{r}
#just a graphics frame
plot = ggplot(data = fulldata, mapping = aes(x = perrep_2012, y = perrep_2016))

#adding a scatterplot layer
plot + geom_point()

#another predictor
ggplot(fulldata, aes(y = perrep_2016, x = median_rent)) + 
    geom_point()

#another predictor
ggplot(fulldata, aes(y = perrep_2016, x = percent_white)) + 
    geom_point()
```
    
    
   
<br>
<br>

### Exercise 7: Side-by-side plots of 1 quantitative variable vs 1 categorical variable


```{r}
#density plots by group
ggplot(fulldata, aes(x = perrep_2016, fill = StateColor)) + 
    geom_density()

#to see better: add transparency
ggplot(fulldata, aes(x = perrep_2016, fill = StateColor)) + 
    geom_density(alpha = 0.5) + 
	scale_fill_manual(values = c("blue", "purple", "red"))

#to see better: split groups into separate plots
ggplot(fulldata, aes(x=perrep_2016, fill=StateColor)) + 
    geom_density(alpha=0.5) + 
    facet_wrap(~StateColor)

#violins instead
ggplot(fulldata, aes(y=perrep_2016, x=StateColor)) + 
    geom_violin()

#boxes instead
ggplot(fulldata, aes(y=perrep_2016, x=StateColor)) + 
    geom_boxplot()

```



<br>
<br>


### Exercise 8: Scatterplots of 1 quantitative variable vs 1 categorical & 1 quantitative variable


```{r}
#scatterplot: id groups using color
ggplot(fulldata, aes(y=perrep_2016, x=percent_white, color=StateColor)) + 
    geom_point(alpha=0.5)

#scatterplot: id groups using shape
ggplot(fulldata, aes(y=perrep_2016, x=percent_white, shape=StateColor)) + 
    geom_point(alpha=0.5)

#scatterplot: split/facet by group
ggplot(fulldata, aes(y=perrep_2016, x=percent_white, color=StateColor)) + 
    geom_point(alpha=0.5) + 
    facet_wrap( ~ StateColor)
```



<br>
<br>


### Exercise 9: Side-by-side plots of 1 quantitative variable vs 2 categorical variables

```{r}
#density plots by group combo
ggplot(fulldata, aes(x=perrep_2016, fill=StateColor)) + 
    geom_density(alpha=0.5) + 
    facet_wrap(~ winrep_2012)

#switch faceting order
ggplot(fulldata, aes(x=perrep_2016, fill=winrep_2012)) + 
    geom_density(alpha=0.5) + 
    facet_wrap(~ StateColor)
```

<br>
<br>

### Exercise 10: Plots of 3 quantitative variables

```{r}
#scatterplot: represent third variable using color
ggplot(fulldata, aes(y=perrep_2016, x=percent_white, color=median_rent)) + 
    geom_point(alpha=0.5)

#scatterplot: discretize the third variable into 2 groups & represent with color
ggplot(fulldata, aes(y=perrep_2016, x=percent_white, color=cut(median_rent,2))) + 
    geom_point(alpha=0.5)

#scatterplot: discretize the third variable into 2 groups & facet by group
ggplot(fulldata, aes(y=perrep_2016, x=percent_white, color=cut(median_rent,2))) + 
    geom_point() + 
    facet_wrap(~ cut(median_rent, 2))
```


    
<br>
<br>


### Exercise 11: Maps!

```{r}
#Construct maps using the choroplethrMaps library

#a map of Trump support


#a map of Trump wins


#a map of percent white


```




<br>
<br>

<br>
<br>



## Linear Regression Models

### Exercise 12: Models with 1 quantitative predictor


```{r}
#visualization

#model

```


<br> 
<br>

### Exercise 13: Models with 1 categorical predictor

```{r}
#plot the relationship

```


```{r}
#detach/unload libraries (a bug)
detach(package:choroplethr)
detach(package:acs)
detach(package:plyr)

#calculate the group means

```



```{r}
#construct the model

```



<br>        
<br>

### Exercise 14: Models with 1 quantitative predictor & 1 categorical predictor


```{r}
#visualize the relationship

#construct the model

```



<br>
<br>


### Exercise 15: Models with 2 quantitative predictors

```{r}
#visualize the relationship

#construct the model


```



<br> 
<br>

### Exercise 16: Models with 2 categorical predictors

```{r}
#create income brackets

```

```{r}
#visualize the relationship

#construct the model

```




<br>
<br>
<br>
<br>


