---
title: "IMA Day 3 Notes"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---




## Getting Started

**Load packages required for exercises:**

```{r}
suppressPackageStartupMessages(library(rvest))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(tidyr))
```


## Scraping & Wrangling Data {-}

**Scrape data from [Wiki](https://en.wikipedia.org/wiki/Current_members_of_the_United_States_House_of_Representatives)**

```{r}
#load the rvest library (if it's not already, install it first!)
suppressPackageStartupMessages(library(rvest))

#copy the Wiki url
url = "https://en.wikipedia.org/wiki/Current_members_of_the_United_States_House_of_Representatives"

#scrape ALL tables from the url
allTables = read_html(url) %>% 
    html_nodes("table") %>% 
    html_table(fill=TRUE)

#the 7th table corresponds to "Voting members by state".  store this as "houseDirty"
houseDirty = allTables[[7]]

#check out the head of houseDirty
head(houseDirty, 3)
```


<br>
<br>

### Exercise 1

```{r}
#part a
mytext <- "This is my text"
gsub("s", "z", mytext)
gsub(" ", "", mytext)
substr(mytext, 1, 3)
substr(mytext, 4, 7)

#part b
mynumber <- "2017"
class(mynumber)
mynewnumber <- as.numeric(mynumber)
mynewnumber + 2
class(mynewnumber)
```

<br>
<br>

### Exercise 2

```{r}
#load the dplyr library for some wrangling functions
suppressPackageStartupMessages(library(dplyr))

#take out the third column (the blank "Party" variable)
house = houseDirty[-3]

#remove the spaces from the variable names
names(house) = gsub(" ", "_", names(house))

#check the class of the Assumedoffice variable 
class(house$Assumed_office)

#in 1 set of code: 
#remove the "*" from the Assumedoffice measurements
#convert Assumedoffice to numeric
house$Assumed_office = as.integer(substr(x = house$Assumed_office, start = 1, stop = 4))

#define an "Age" variable
house = mutate(.data = house, Age = as.integer(2017) - Born_in)

#confirm that your house is consistent with the results in the notes
dim(house)
head(house[,c(1,6,7,8)])
mean(house$Assumed_office)
```

<br>
<br>

## Sampling Variability
```{r}
suppressPackageStartupMessages(library(ggplot2))

mu = mean(house$Age, na.rm=TRUE)

sigma = sd(house$Age, na.rm=TRUE)

ggplot(house, aes(x=Age)) + 
    geom_histogram(color="white") + 
    labs(title="POPULATION distribution")
```


### Exercise 3

```{r}
#choose your own random number seed
set.seed(65536)

#take a sample of 16 members
sample1 = sample_n(house, size=16)

#check the properties
head(sample1)
dim(sample1)
```


part b:
```{r}
ggplot(sample1, aes(x=Age)) + 
    geom_histogram(color="white", binwidth=5) + 
    labs(title="SAMPLE Distribution (n=16)")
```

part c:
```{r}
mu_est_1 = mean(sample1$Age)
sigma_est_1 = sd(sample1$Age)
```




<br>
<br>

### Exercise 4

```{r}
#take another sample
sample2 <- sample_n(house, size=16)

head(sample2)

mu_est_2 = mean(sample2$Age)

sigma_est_2 = sd(sample2$Age)
```




<br>
<br>


## Sampling Distributions



<br>

Sample Size   Shape         Min $\overline{y}$  Mean $\overline{y}$   Max $\overline{y}$  St Dev of $\overline{y}$ 
------------- ------------- ------------------- --------------------- ------------------- --------------------------
$n = 16$
$n = 64$
$n = 100$



<br>
<br>

### Exercise 5

```{r}
#set the seed
set.seed(2)

#initialize the for loop
simulation <- data.frame(n16=rep(0,1000), n64=rep(0,1000), n100=rep(0,1000))

#run the for loop
for(i in 1:1000){
	#take a sample of 16 members
	s16 = sample_n(house, size=16)
	
	#calculate the mean age & store in simulation$n16
	simulation$n16[i] = mean(s16$Age)
	
	#take a sample of 64 members
	s64 = sample_n(house, size=64)
	
	#calculate the mean age & store in simulation$n64
	simulation$n64[i] = mean(s64$Age)
	
	#take a sample of 100 members
	s100 = sample_n(house, size=100)
	
	#calculate the mean age & store in simulation$n100
	simulation$n100[i] = mean(s100$Age)
}
```


<br>
<br>

### Exercise 6

```{r}
ggplot(simulation, aes(x=n16)) + 
    geom_histogram(color="white", binwidth=2) + 
    labs(title="SAMPLING Distribution") + 
    lims(x=c(45,75)) + 
    geom_vline(xintercept=58.54734, color="red")
```


    


<br>
<br>

### Exercise 7

```{r}
ggplot(simulation, aes(x=n64)) + 
    geom_histogram(color="white", binwidth=2) + 
    labs(title="SAMPLING Distribution") + 
    lims(x=c(45,75)) + 
    geom_vline(xintercept=58.54734, color="red")
```


<br>
<br>

### Exercise 8

```{r}
ggplot(simulation, aes(x=n100)) + 
    geom_histogram(color="white", binwidth=2) + 
    labs(title="SAMPLING Distribution") + 
    lims(x=c(45,75)) + 
    geom_vline(xintercept=58.54734, color="red")
```


<br>
<br>

### Exercise 9

```{r}
#follow directions

```


<br>
<br>

### Exercise 10

```{r}
#density plot by group

```


<br>
<br>

### Exercise 11





<br>
<br>


## Regression Models

### Exercise 12

part a:
```{r}
#set the seed

#initialize the for loop


#run the for loop

```
    
part b:

```{r}
#construct a histogram of the 1000 beta0 estimates

#construct a histogram of the 1000 beta1 estimates

```

part c:
```{r}
#approximate the standard errors

```


<br>
<br>

### Exercise 13

part a:
```{r}

```

part b:
```{r}

```


    
    

    
    

<br>
<br>


## The Central Limit Theorem

### Exercise 14



<br>
<br>


## Extra: Writing Functions






<br>
<br>
<br>


