---
title: "IMA Day 2 Notes"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
---




## Getting Started

**Load packages required for exercises:**

```{r}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(mosaic))
```




<br>
<br>



## Regression Assumptions & Residual Analysis 

### Exercise 1




<br>
<br>

### Exercise 2


```{r}
suppressPackageStartupMessages(library(mosaic))
data(Galton)
ggplot(Galton, aes(y=height, x=father)) + 
    geom_point() + 
    geom_smooth(method="lm")
GaltonMod <- lm(height ~ father, data=Galton)
summary(GaltonMod)
```



```{r}
#combine the raw responses, model predictions, and model residuals
ModResults = data.frame(observed = Galton$height, predicted = GaltonMod$fitted.values, residual = GaltonMod$residuals)
head(ModResults)

#a plot of residuals versus the predictions    
ggplot(ModResults, aes(y=residual, x=predicted)) + 
    geom_point() + 
    geom_hline(yintercept=0)

#a Q-Q plot of the residuals
ggplot(ModResults, aes(sample=residual)) + 
    geom_qq()
```




<br>
<br>
<br>
<br>


## Measuring model quality: $R^2$

<br>


```{r}
fulldata <- read.csv("https://www.macalester.edu/~ajohns24/data/IMAdata1.csv")

# add a variable
fulldata <- mutate(.data = fulldata, beforeL = (as.character(county) < "L"))

#set the random number seed
set.seed(2000)

#load the dplyr package for the sample_n function
suppressPackageStartupMessages(library(dplyr))

#take a sample
subdata <- sample_n(fulldata, size=100)
dim(subdata)

```



<br>

### Exercise 3


`mod1`:

```{r}
#fit the model
mod1 = lm(perrep_2016 ~ perrep_2012, data = subdata)

#store the results
mod1results =  data.frame(observed = subdata$perrep_2016[-mod1$na.action], predicted = mod1$fitted, residual = mod1$resid)

#check out the head
head(mod1results,3)

#variance calculations
var(mod1results$observed)
var(mod1results$predicted)
var(mod1results$residual)
```
   

<br>
<br>

`mod2`:

```{r}
#fit the model
mod2 = lm(perrep_2016 ~ median_rent, data = subdata)

#store the results
mod2results =  data.frame(observed = subdata$perrep_2016[-mod2$na.action], predicted = mod2$fitted, residual = mod2$resid)

#check out the head
head(mod2results,3)

#variance calculations
var(mod2results$observed)
var(mod2results$predicted)
var(mod2results$residual)
``` 



<br>
<br>

`mod3`:

```{r}
#fit the model
mod3 = lm(perrep_2016 ~ beforeL, data = subdata)

#store the results
mod3results =  data.frame(observed = subdata$perrep_2016[-mod3$na.action], predicted = mod3$fitted, residual = mod3$resid)

#check out the head
head(mod3results,3)

#variance calculations
var(mod3results$observed)
var(mod3results$predicted)
var(mod3results$residual)
``` 

    

<br>
<br>

### Exercise 4

Model   Predictor       Var(response)   Var(predictions)    Var(residuals)  $R^2$
------- --------------- --------------- ------------------- --------------- -----------
`mod1`  `perrep_2012`   0.02690597      0.02443953          0.002466443
`mod2`  `median_rent`   0.02690597      0.01023033          0.01667564
`mod3`  `beforeL`       0.02690597      0.001113426         0.02579254
    

<br>
<br>

### Exercise 5


    
<br>
<br>
<br>
<br>




## Measuring model quality: cross validation

### Exercise 6

```{r}
#take a sample of size 1 (try this a few times)

```

```{r}
#set the seed

#take a sample of size 1 (try this a few times)

```




<br>
<br>

### Exercise 7

```{r}
#IMPORTANT: set the random number seed

#sample 50 of the 100 cases for the training set


#take the the other 50 for testing


#check out the dimensions


```

```{r}
#fit the model

#calculate MSE


#set up prediction function


#make predictions


#calculate residuals


#calculate MSE


#plot


```





<br>
<br

### Exercise 8 



<br>
<br>


### Exercise 9

```{r}

```




<br>
<br>


### Exercise 10


```{r}
#initialize the loop: set up a storage location for the 10 squared elements

#loop through all steps btwn brackets 10 times

#check out the results

```



<br>
<br>


### Exercise 11


```{r}
#initialize the loop: set up a storage location for the 10 squared elements

#loop through all steps btwn brackets 10 times

#check out the results

```


<br>
<br>
<br>
<br>


