# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data
```{r Initialization}
unzip("activity.zip")
activity = read.csv(file='activity.csv')
```



## What is mean total number of steps taken per day?
```{r Interday, results='asis'}
#	Daily activity rollup
activityInterday = aggregate(x=activity$steps,
							 by=list(date = activity$date),
							 FUN=sum)
names(activityInterday)[2] = "steps"

#	Plot
hist(activityInterday$steps, breaks=64, density=64,
	 main="Total number of steps taken per day")

#	Summary statistics
meanActivityInterday   = mean(  activityInterday$steps, na.rm=TRUE)
medianActivityInterday = median(activityInterday$steps, na.rm=TRUE)
cat("Mean  \t=", meanActivityInterday,   "\n")
cat("Median\t=", medianActivityInterday, "\n")
```



## What is the average daily activity pattern?
```{r Intraday, results='asis'}
#	Compute
activityIntraday = aggregate(x=activity$steps,
							 by=list(interval=activity$interval),
							 FUN=function(x) mean(x, na.rm=TRUE))
names(activityIntraday)[2] = "steps"

#	Display
plot(activityIntraday, type="l",
	 main="Mean activity throught the day")
cat("Interval having the maximum number of steps =",
	which.max(activityIntraday$steps), "\n")
```



## Imputing missing values
To replace missing values, we use the mean number of steps over all the days for that particular time interval.
```{r Imputing, results='asis'}
#	Total rows of NAs
cat("Total number of missing rows =", sum(!complete.cases(activity)), "\n")

#	Imputation using the mean of the 5 minute interval
naIdx = !complete.cases(activity)
activityImputed = activity
activityImputed[naIdx, "steps"] = activityIntraday[
	which(activityIntraday$interval == activityImputed$interval[naIdx]), "steps"]

#	Aggregation
activityInterdayImputed = aggregate(
	x=activityImputed$steps,
	by=list(date = activityImputed$date),
	FUN=sum)
names(activityInterdayImputed)[2] = "steps"

#	Histogram
hist(activityInterdayImputed$steps, breaks=64, density=64,
	 main="Total number of steps taken per day")

#	Summary statistics
meanActivityInterdayImputed   = mean(  activityInterdayImputed$steps, na.rm=TRUE)
medianActivityInterdayImputed = median(activityInterdayImputed$steps, na.rm=TRUE)
cat("Mean after imputation  \t=", meanActivityInterdayImputed,   "\n")
cat("Median after imputation\t=", medianActivityInterdayImputed, "\n")

#	Comparison
if (meanActivityInterday != meanActivityInterdayImputed) {
	cat("The values of the mean for the imputed data is different.\n")
}  else {
	cat("The values of the mean for the imputed data are the same.\n")
}

if (medianActivityInterday != medianActivityInterdayImputed) {
	cat("The values of the median for the imputed data is different.\n")
}  else {
	cat("The values of the median for the imputed data are the same.\n")
}

```



## Are there differences in activity patterns between weekdays and weekends?
```{r WeekdayWeekend, results='asis'}
day = weekdays(as.Date(activityImputed$date), abbreviate=TRUE)
daytype = sapply(X=day, FUN=function(d) {
	ifelse(d == "Sat" || d == "Sun", "weekend", "weekday")
})
names(daytype) = NULL
daytype = as.factor(daytype)
activityImputed$daytype = daytype

#	Aggregation
activityImputedIntraday = aggregate(
	x=activityImputed$steps,
	by=list(interval=activityImputed$interval, daytype=activityImputed$daytype),
	FUN=function(x) mean(x, na.rm=TRUE))
names(activityImputedIntraday)[3] = "steps"

#	Plotting
par(mfrow=c(2,1))
plot(x = activityImputedIntraday[, c(1,3)]
	[which(activityImputedIntraday$daytype == "weekend"),],
	 type="l", main="Weekend")
plot(x = activityImputedIntraday[, c(1,3)]
	 [which(activityImputedIntraday$daytype == "weekday"),],
	 type="l", main="Weekday")
```
